Variational Monte Carlo on Bosonic Systems

### 2018-04-05 project_1/article/article.tex
Spell check section on scaling the system
Spell check MH algorithm section

    To determine if we have located a minimum we can use the variance of the
    variational energy.
    \begin{align}
        \sigma^2
        &=
        \frac{
            \bra{\Psi_T}H^2\ket{\Psi_T}
        }{
            \bra{\Psi_T}\ket{\Psi_T}
        }
        - \para{
            \frac{
                \bra{\Psi_T}H\ket{\Psi_T}
            }{
                \bra{\Psi_T}\ket{\Psi_T}
            }
        }^2.
    \end{align}
    If we have found the ``true'' trial wavefunction $\ket{\Psi_T}$ with the
    correct variational parameters, i.e., the eigenfunction to the Hamiltonian
    we get $\sigma^2 = 0$. We are thus interested in trying to find the
    variational parameters in the trial wavefunction that minimizes the
    variance.

@@

        \subsubsection{Scaling the system}
            We now introduce a scaled distance $\vf{r}' = \vf{r}/a_{\text{ho}}$
            \cite{dubois2001bose}, where
            \begin{align}
                a_{\text{ho}}
                &=
                \sqrt{\frac{\hbar}{m\omega}}.
            \end{align}
            We can then rewrite the Hamiltonian for the elliptic potential in terms
            of this new distance. By doing a variable substitution for each
            direction in the Laplace operator we get
            \begin{align}
                \nabla_k^2
                &=
                \frac{1}{a_{\text{ho}}^2}{\nabla_k'}^2.
            \end{align}

            Looking at the one-body part of the Hamiltonian we get
            \begin{align}
                H
                &=
                \sum_{k = 1}^N
                \brac{
                    -\frac{\hbar^2}{2m}\nabla_k^2
                    + \half m\brak{
                        \omega^2(x_k^2 + y_k^2)
                        + \omega_z^2 z_k^2
                    }
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }
                \equiv
                \phi(\vf{r}_k').
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.

@@

        \begin{enumerate}
            \item The system is initialised by a certain number $N$ of randomly
                generated positions, or particles. This allows us to evaluate
                the wavefunction at these points and compute the local energy
                $E_L$.

@@

            \item A ratio between new wavefunction density and the previous
                (initial) density is computed and compared to a random number.
                This acceptance probability decides if the particle move is
                rejected or accepted.  The particle is only allowed to move a
                predetermined step length.

            \item If the particle movement is accepted and the local energy
                $E_L$ is computed for the new system.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a transition
        probability and the acceptance probability.  The strength of the
        algorithm is that the transition algorithm need not be known. For
        example, the simplest case is to accept the new state, i.e., the new
        position for the random walker, if the ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1)$.

@@

            \subsubsection{Stability of the algorithm}
                When employing the method of gradient descent to locate a
                minimum value for the variational parameters we expect some
                issues due to stability of the Monte Carlo methods and the
                steepness of the expected energy curves. If our guess of the
                location of the minimum is far off we can expect the method to
                diverge as the gradient will become huge. This will yield a
                new value for $\vfg{\alpha}$ in
                \autoref{eq:parameter_difference} which will be further away and
                might result in a local minimum ``trapping'' the method. To
                prevent this from happening we have added a limit to the allowed
                change in $\vfg{\alpha}$. If the change in $\vfg{\alpha}$ is
                larger than a set order of magnitude we keep dividing
                $\vfg{\alpha}$ with this number until this criteria is
                fulfilled.

@@

        We will start without importance sampling, that is by employing the
        simpler and more naïve ``brute force'' Monte Carlo sampling. Moreover, we
        will compute the value for the local energy analytically as well as with
        a numerical approach.  In the numerical approach we employ the well-know
        approximation for the second derivative,
        \begin{equation}
                f''(x) \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
        \end{equation}

@@

        \subsubsection{Finding the minimum}
            Using the method of gradient descent we choose a set of six starting
            $\alpha_0 \in [0.2, 0.8]$ and try to locate a value for $\alpha$
            where the gradient goes to zero. We looked at $N = 10$ with $d = 3$,
            $\omega = 1$ and $\beta = \lambda = \sqrt{8}$. This yields the value
            for the minimum $\alpha$ to be
            \begin{align}
                \alpha = 0.49744 \pm 0.00002.
                \label{eq:optimal_interacting_alpha}
            \end{align}
            In \autoref{fig:gradient_descent_interacting} we can see how the
            conjugate gradient method ``moves'' to find the optimal value of
            $\alpha$.

@@

        \subsubsection{One-Body Densities}
            We now look at one-body densities for the value of $\alpha$ found in
            \autoref{eq:optimal_interacting_alpha}. The results for a system consisting
            of $n=100$ particles are displayed in \autoref{fig:one_body_density_100}.
            In this figure we see the densities for for a system confined to a
            perturbed elliptic trap with and without interaction. We see that for the
            interacting case, the one-body density is skewed towards a larger
            $\abs{\vf{r}}$, as is expected. The ``ideal'' density, a system with spherical
            trap, is also included in the figure.

@@

            \begin{figure}
                \includegraphics[width=244px]
                {../data/figures/problem_g_100N.pdf}
                \caption{In this figure the interacting and non-interacting
                densities correspond to the elliptical potential with and
                without the Jastrow factor respectively. The ``ideal'' system is
                the spherical potential with the non-interacting Gaussian trial
                wavefunction.}
                \label{fig:one_body_density_100}
            \end{figure}

@@

    \subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new ``particle'' and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions.}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \emph{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of the former with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.

>>>>

    To determine if we have located a minimum we can use the variance of the
    variational energy.
    \begin{align}
        \sigma^2
        &=
        \frac{
            \bra{\Psi_T}H^2\ket{\Psi_T}
        }{
            \bra{\Psi_T}\ket{\Psi_T}
        }
        - \para{
            \frac{
                \bra{\Psi_T}H\ket{\Psi_T}
            }{
                \bra{\Psi_T}\ket{\Psi_T}
            }
        }^2.
    \end{align}
    If we have found the "true" trial wavefunction $\ket{\Psi_T}$ with the
    correct variational parameters, i.e., the eigenfunction to the Hamiltonian
    we get $\sigma^2 = 0$. We are thus interested in trying to find the
    variational parameters in the trial wavefunction that minimizes the
    variance.

@@

        \subsubsection{Scaling the system}
            We now introduce a scaled distance $\vf{r}' = \vf{r}/a_{\text{ho}}$
            \cite{dubois2001bose}, where
            \begin{align}
                a_{\text{ho}}
                &=
                \sqrt{\frac{\hbar}{m\omega}}.
            \end{align}
            We can then rewrite the Hamiltonian for the elliptic potential in terms
            of this new distance. By doing a variable substitution for each
            direction in the Laplace operator we get
            \begin{align}
                \nabla_k^2
                &=
                \frac{1}{a_{\text{ho}}^2}{\nabla_k'}^2.
            \end{align}
            The one-body part of the Hamiltonian then yields
            \begin{align}
                H
                &=
                \sum_{k = 1}^N
                \brac{
                    -\frac{\hbar^2}{2m}\nabla_k^2
                    + \half m\brak{
                        \omega^2(x_k^2 + y_k^2)
                        + \omega_z^2 z_k^2
                    }
                }
                \\
                &=
                \frac{\hbar\omega}{2}
                \sum_{k = 1}^N
                \brac{
                    -{\nabla_k'}^2
                    +
                    \brak{
                        {x_k'}^2 + {y_k'}^2 + \lambda^2{z_k'}^2
                    }
                },
                \label{eq:scaled_hamiltonian}
            \end{align}
            where we have introduced the dimensionless frequency $\lambda =
            \omega_z/\omega$. The single particle functions also get a scaling
            factor
            \begin{align}
                \phi_k
                &=
                \exp\brak{
                    -\frac{\alpha\hbar}{m\omega}
                    ({x_k'}^2 + {y_k'}^2 + \beta^2{z_k'}^2)
                }.
                \label{eq:scaled_spf}
            \end{align}
            The correlation wavefunction and the interaction potential remain the
            same, but using the scaled hard core radius $a/a_{\text{ho}} =
            0.0043$.

@@

        \begin{enumerate}
            \item The system is initialised by a certain number $N$ of randomly
                generated positions, $\vf{r} = (\vf{r}_1, \dots, \vf{r}_N)$,
                which represent the particles. This allows us to evaluate the
                wavefunction at these points and compute the local energy
                $E_L(\vf{r})$.

@@

            \item A ratio, $q$, between the probability density of the new
                wavefunction and the previous probability density is computed.
                We accept the step if $q \geq 1$, else we accept the step with a
                probability $p$. The latter is done by drawing a random uniform
                number, $x \in [0, 1]$, and checking if $q > x$.  This
                acceptance probability decides if the particle move is rejected
                or accepted.  The particle is only allowed to move a
                predetermined step length.

            \item If the particle movement is accepted then the local energy is
                computed for the new system.

@@

        The algorithm described above can be applied in an "exhaustive" search
        of the parameter space in order to find the optimal parameters.  Whether
        a proposed move is accepted or not is determined by a transition
        probability and the acceptance probability.  The strength of the
        algorithm lies in that the transition algorithm need not be known. For
        example, the simplest case is to accept the new state, i.e., the new
        position for the random walker, if the ratio
        \begin{align}
            q(\vf{r}_{i + 1}, \vf{r}_i)
            &=
            \frac{\left|\Psi_T(\vf{r}_{i + 1})\right|^2}
            {\left|\Psi_T(\vf{r}_{i})\right|^2},
        \end{align}
        where $\vf{r}_{i + 1}$ are all the positions at step $i + 1$, is greater
        than a uniform probability $p \in [0, 1]$.

@@

            \subsubsection{Stability of the algorithm}
                When employing the method of gradient descent to locate a
                minimum value for the variational parameters we expect some
                issues due to stability of the Monte Carlo methods and the
                steepness of the expected energy curves. If our guess of the
                location of the minimum is far off we can expect the method to
                diverge as the gradient will become huge. This will yield a
                new value for $\vfg{\alpha}$ in
                \autoref{eq:parameter_difference} which will be further away and
                might result in a local minimum "trapping" the method. To
                prevent this from happening we have added a limit to the allowed
                change in $\vfg{\alpha}$. If the change in $\vfg{\alpha}$ is
                larger than a set order of magnitude we keep dividing
                $\vfg{\alpha}$ with this number until this criteria is
                fulfilled.

@@

        We will start without importance sampling, that is by employing the
        simpler and more naïve "brute force" Monte Carlo sampling. Moreover, we
        will compute the value for the local energy analytically as well as with
        a numerical approach.  In the numerical approach we employ the well-know
        approximation for the second derivative,
        \begin{equation}
                f''(x) \approx \frac{f(x-h) - 2f(x) + f(x+h)}{h^2}.
        \end{equation}

@@

        \subsubsection{Finding the minimum}
            Using the method of gradient descent we choose a set of six starting
            $\alpha_0 \in [0.2, 0.8]$ and try to locate a value for $\alpha$
            where the gradient goes to zero. We looked at $N = 10$ with $d = 3$,
            $\omega = 1$ and $\beta = \lambda = \sqrt{8}$. This yields the value
            for the minimum $\alpha$ to be
            \begin{align}
                \alpha = 0.49744 \pm 0.00002.
                \label{eq:optimal_interacting_alpha}
            \end{align}
            In \autoref{fig:gradient_descent_interacting} we can see how the
            conjugate gradient method "moves" to find the optimal value of
            $\alpha$.

@@

        \subsubsection{One-Body Densities}
            We now look at one-body densities for the value of $\alpha$ found in
            \autoref{eq:optimal_interacting_alpha}. The results for a system consisting
            of $n=100$ particles are displayed in \autoref{fig:one_body_density_100}.
            In this figure we see the densities for for a system confined to a
            perturbed elliptic trap with and without interaction. We see that for the
            interacting case, the one-body density is skewed towards a larger
            $\abs{\vf{r}}$, as is expected. The "ideal" density, a system with spherical
            trap, is also included in the figure.

@@

            \begin{figure}
                \includegraphics[width=244px]
                {../data/figures/problem_g_100N.pdf}
                \caption{In this figure the interacting and non-interacting
                densities correspond to the elliptical potential with and
                without the Jastrow factor respectively. The "ideal" system is
                the spherical potential with the non-interacting Gaussian trial
                wavefunction.}
                \label{fig:one_body_density_100}
            \end{figure}

@@

    \subsection{Energy per particle}
        For the non-interacting harmonic oscillator system there is no
        difference between a new particle and new dimensions
        \footnote{This is not entirely true in the case of our implementation of
        the Metropolis algorithms as we draw a single random particle and move
        all dimensions instead of iterating over all particles and all
        dimensions.}. This means that the inclusion of a new particle or
        dimension yields a linear scaling in the energy. Thus if we compute the
        energy per particle for systems of varying size we expect to get the
        same energy. This is most easily seen in \autoref{fig:initial_problem_b}
        and in the expression for the exact energy \autoref{eq:exact_energy}.
        This, however, is \emph{not} the case in the interacting system due to
        the Jastrow factor. Looking at \autoref{tab:10_interacting} and
        \autoref{tab:100_interacting} we see that multiplying the expected
        energy of the former with $10$ yields a lower number than the latter. We
        can also see this behavior in \autoref{fig:interacting}, where the
        energy per particle for $N = \{10, 50, 100\}$ have been plotted in the
        same figure.
